---
title: "Responsible GenAI for Coding in R"
author: "Laurence Frank"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '3'
subtitle: Building a grouped boxplot with the BRAVE(R) framework
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4.5)
```

# Purpose of the exercise

In this exercise you will learn how to use an AI tool to assist you with writing
code in R using the BRAVE(R) framework and keeping human oversight. This is in 
contrast with letting the AI tool write the enitire code from scratch 
(*ghostwriting*): the AI produces R code to build a chart, but you do not learn 
how it works, and you cannot tell whether it is right.

Instead you will proceed as follows:

1. **Find reliable code** from a trusted, curated source.
2. **Understand it** line by line, using AI as a *tutor*, not an author.
3. **Adapt it** to your own study design.
4. **Verify** the result (human oversight).

During the exercise, AI is a sparring partner that *explains* and *suggests*. You
remain the author and the final reviewer. If you cannot evaluate the output, that
is a signal that you need to understand the topic better before you rely on the
tool (human oversight).

> **Data safety reminder.** Never paste private, unpublished, participant, or
> otherwise sensitive data into an AI tool. That is why this exercise
> uses **simulated data** that you generate yourself. The simulation code is
> provided below.

## Learning objectives

By the end of this exercise you will be able to:

- Locate trustworthy graphics code in the **R Graph Gallery**.
- Read and **annotate** ggplot2 code so that every line has a comment in your own words.
- Adapt a grouped boxplot from a **7 × 2** to a **4 × 3** study design.
- Write prompts using the **BRAVE(R)** framework (Boundaries, Role, Audience,
  Variables, Expectations, Refine) to make Generative AI *explain* and *refine* code.
- Apply **human-oversight checks** to confirm the boxplot matches the data.

---

# Set up your R environment

You only need to do **one** of the two options below.

### Option A: You already have RStudio installed

1. Open **RStudio**.
2. Create a new R Markdown file (or just open *this* file):
   `File ▸ New File ▸ R Markdown…`, then keep the HTML output and click **OK**.
   To open this handout: `File ▸ Open File…` and select `R_exercise.Rmd`.
3. Install the one package you need (run this **once** in the **Console**, not in a chunk):

   ```r
   install.packages("ggplot2")
   ```

4. To run a code chunk, click the small green ▶ ("Run Current Chunk") at the
   top-right of the chunk, or place your cursor in it and press
   **Ctrl+Shift+Enter** (Windows/Linux) / **Cmd+Shift+Enter** (Mac).
5. To produce the full HTML document, click **Knit** at the top of the editor.

### Option B: You do not have RStudio: use Posit Cloud (in your browser)

[Posit Cloud](https://posit.cloud) runs RStudio in the browser — nothing to install.

1. Go to **https://posit.cloud** and click **Sign Up** (the free plan is enough).
2. After logging in, click **New Project ▸ New RStudio Project**. Wait a few
   seconds for the workspace to load. You now have the full RStudio interface.
3. Upload this file: in the **Files** pane (bottom-right) click **Upload**,
   choose `R_exercise.Rmd`, then click it to open.
4. Install the package once in the **Console**:

   ```r
   install.packages("ggplot2")
   ```

5. Run chunks with the green ▶ button and **Knit** exactly as in Option A.

**Tip.** On the free Posit Cloud plan, projects have limited monthly hours.
Close the project tab when you are done so you don't burn through your time.

### Load the package

Once `ggplot2` is installed, load it. This chunk runs at the top of every session.

```{r load-packages}
library(ggplot2)
```

---

# Step 1: Study the reference graph

Open the R Graph Gallery page for a grouped boxplot:

**https://r-graph-gallery.com/265-grouped-boxplot-with-ggplot2.html**

The gallery example visualises **7 groups** (varieties A–G) split into **2
subgroups** (`high` / `low`) — a **7 × 2** design. The reference code builds its
data and plot like this:

```{r reference-gallery, fig.height=4}
# --- Reference code from the R Graph Gallery (7 x 2 design) ---
# (Reproduced here only so you can compare it to your own version.)

variety   <- rep(LETTERS[1:7], each = 40)
treatment <- rep(c("high", "low"), each = 20)
note      <- seq(1:280) + sample(1:150, 280, replace = TRUE)
gallery_data <- data.frame(variety, treatment, note)

ggplot(gallery_data, aes(x = variety, y = note, fill = treatment)) +
  geom_boxplot()
```

Notice the two ideas that make a *grouped* boxplot work:

- the **group** factor goes on the `x` axis (`x = variety`);
- the **subgroup** factor goes in `fill = treatment`, which splits each x
  position into one box per subgroup.

---

# Step 2: Your scenario and simulated data

**The situation.** You want a grouped boxplot for your data, like the gallery example, but the
gallery shows a 7 × 2 design and **your** study is **4 × 3**. Suppose you have collected data from childeren, adolescents and adults about wellbeing in four different regions.

- **Factor 1 = `region`** (4 levels): `North`, `South`, `East`, `West`
- **Factor 2 = `age_group`** (3 levels): `Children`, `Adolescents`, `Adults`
- **Response variable = `wellbeing`**: a continuous self-report index (higher = better).
- 20 respondents per combination leads to 4 × 3 × 20 = **240 observations**.

Run the chunk below to create the simulated dataset. The simulation is fully provided, is
reproducible (`set.seed`), and uses no external files.

```{r simulate-data}
set.seed(123)  # makes the random draw reproducible: everyone gets the same data

n_rep <- 20    # number of respondents per region x age_group combination

# Build every combination of respondent, age_group and region.
# expand.grid() varies the FIRST listed factor fastest.
study <- expand.grid(
  respondent = 1:n_rep,
  age_group  = c("Children", "Adolescents", "Adults"),
  region     = c("North", "South", "East", "West")
)

# Effect sizes we are "baking into" the simulation, so we know the right answer.
region_effect <- c(North = 0, South = 3, East = -2, West = 1)
age_effect    <- c(Children = 0, Adolescents = -4, Adults = 2)  # an "adolescent dip"

# Build the response: baseline 65 + region effect + age effect + noise.

study$wellbeing <- 65 +
  region_effect[as.character(study$region)] +
  age_effect[as.character(study$age_group)] +
  rnorm(nrow(study), mean = 0, sd = 6)

# Fix the order in which levels appear on the plot (instead of alphabetical).
study$region    <- factor(study$region,
                          levels = c("North", "South", "East", "West"))
study$age_group <- factor(study$age_group,
                          levels = c("Children", "Adolescents", "Adults"))
```

---

# Step 3 — Understand the code *before* you change it

Here you use AI as a **tutor** to explain the reference code so that you understand every
line. Use the **BRAVE(R)** framework to write a focused prompt.

### A BRAVE(R) prompt to explain the reference code

The prompt might look like this:

*Please explain, line by line, how the following R code builds a grouped
boxplot **(E – Expectations)**. For this task, take on the role of an R coding
tutor **(R – Role)**. The explanation should suit a researcher
who knows basic R but is new to ggplot2 **(A – Audience)**. For each line, tell
me what it does and why it is needed, and explain specifically what the
aes(), fill, and geom_boxplot() parts do **(V – Variables)**. Keep it
concise: one or two sentences per line, no rewritten code **(B – Boundaries)**.*

### Copy the following prompt in an AI tool together with the R code:

Please explain, line by line, how the following R code builds a grouped
boxplot. For this task, take on the role of an R coding tutor. 
The explanation should suit a researcher who knows basic R but is new to ggplot2.
For each line, tell me what it does and why it is needed, and explain specifically 
what the `aes()`, `fill`, and `geom_boxplot()` parts do. Keep it
concise: one or two sentences per line, no rewritten code.

ggplot(gallery_data, aes(x = variety, y = note, fill = treatment)) +
  geom_boxplot()



### Verify
Read the explanation against the official ggplot2 documentation (`?ggplot`, 
`?aes`, `?geom_boxplot` in the Console, or https://ggplot2.tidyverse.org). 
If the explanation and the docs disagree, the docs win.

### Refine

If a part is still unclear, follow up. This might be a Refine step of the 
BRAVE(R) framework. Copy and past this prompt into the AI tool:

_I still don't understand the difference between mapping a variable to
fill versus to x. Can you explain what changes in the plot, with a one-line
example for each?_

---

# Step 4: Adapt and build your plot

Remember the changes you want to make in the plot: 

| Reference (7 × 2) | Your design (4 × 3) |
|---|---|
| `x = variety` (7 groups) | `x = region` (4 groups) |
| `fill = treatment` (2 subgroups) | `fill = age_group` (3 subgroups) |
| `y = note` | `y = wellbeing` |
| `data = gallery_data` | `data = study` |

### A BRAVE(R) prompt to help you adapt the code

If you get stuck, ask AI to *guide* the change, not to hand you a finished
script by using this example prompt:

_I have a working grouped boxplot for a 7 × 2 design and a simulated dataset
`study` for a **4 × 3** design with columns `region` (4 levels), `age_group`
(3 levels) and `wellbeing` **(V)**. I expect to adapt the code myself, so please
tell me which arguments I need to change and why, rather than giving me the
full solution **(E, B)**. Act as an R tutor **(R)** writing for a social-science
researcher new to ggplot2 **(A)**. This is the code that I have now:_

ggplot(gallery_data, aes(x = variety, y = note, fill = treatment)) +
  geom_boxplot()

### Copy past your prompt in the AI tool and ajdust the code

I have a working grouped boxplot for a 7 × 2 design and a simulated dataset
`study` for a **4 × 3** design with columns `region` (4 levels), `age_group`
(3 levels) and `wellbeing`. I expect to adapt the code myself, so please
tell me which arguments I need to change and why, rather than giving me the
full solution. Act as an R tutor writing for a social-science
researcher new to ggplot2. This is the code that I have now:_

ggplot(gallery_data, aes(x = variety, y = note, fill = treatment)) +
  geom_boxplot()


```{r build-plot-task, eval=FALSE}

# >>> YOUR TASK: complete the grouped boxplot for the 4 x 3 design <<<


ggplot(___, aes(x = ____, y = ____, fill = ____)) +
  geom_boxplot()
```


If you have adjusted the code correctly, the plot should look like this:

```{r solutionplot, echo=FALSE}
ggplot(study, aes(x = region, y = wellbeing, fill = age_group)) +
  geom_boxplot()
```
---

# Step 5: Human oversight, check the output

A plot is only useful if it faithfully represents the data. Before you trust it,
run these checks yourself:

- **Count the boxes.** You should see 4 regions × 3 age groups = 12 boxes
  (4 x-axis positions, each split into 3 coloured boxes). If you see a different
  number, the mapping is wrong.
- **Check the legend.** It should list exactly the 3 age groups.
- **Sanity-check against what you simulated.** You built in an *adolescent dip*
  (`Adolescents` lowest, `Adults` highest) and made `South` the highest region.
  Do the box positions broadly match that? If the chart contradicts the known
  truth, suspect the *code*, not the data.

Confirm your counts numerically. This is your "ground truth" to compare the plot against:

```{r oversight-check}
# How many boxes should appear, and is the design balanced?

nlevels(study$region) * nlevels(study$age_group)  # expected number of boxes
table(study$region, study$age_group)              # all cells should be 20

# Median wellbeing per cell — the plot's boxes should track these numbers.
aggregate(wellbeing ~ region + age_group, data = study, FUN = median)
```

---

# Step 6 — Refine the plot

Once the core plot is correct, you can refine its appearance. Treat each
refinement as a small change that can be reviewed. Add one layer, re-knit, look. 
A BRAVE(R) refine prompt might look like this:

My grouped boxplot is correct and I now want to polish it **(E)**. As an R/ggplot2
tutor **(R)**, for a researcher new to ggplot2 **(A)**, suggest how
to add a clear axis labels and a title, apply theme_minimal(), and choose
colour-blind-friendly colours **(V)**. Show me what I need to add to the R code 
below and add comments to each new line of code explaining what it does and 
why it is necessary to add **(B)**.

### Copy and past the prompt into the AI tool with the R code

My grouped boxplot is correct (see R code below) and I now want to polish it. 
As an R/ggplot2 tutor, for a researcher new to ggplot2, suggest how
to add a clear axis labels and a title and choose colour-blind-friendly colours. 
Show me what I need to add to the R code below and add comments to each new line 
of code explaining what it does and why it is necessary to add.

ggplot(study, aes(x = region, y = wellbeing, fill = age_group)) +
  geom_boxplot()
  

A refined version would look like this:

```{r refine-example}

ggplot(study, aes(x = region, y = wellbeing, fill = age_group)) +
  geom_boxplot() +
  labs(
    title = "Wellbeing by region and age group (4 x 3 design)",
    x = "Region",
    y = "Wellbeing index",
    fill = "Age group"
  ) +
  scale_fill_viridis_d()   
```

---

# Step 7 — Reflection

Write a few sentences in answer to each:

1. Which BRAVE(R) component made the biggest difference to the usefulness of the
   AI's answers, and why?
2. Where did human oversight catch something the AI got wrong or that you simply 
   needed to verify?
3. How is "find reliable code and understand it" different from "ask the AI to
   write it for me" in terms of the reliability of your results?
4. The simulation uses `set.seed(123)`. Why does reproducibility matter when you
   are checking whether your code is correct?
---

*Reference graph: R Graph Gallery, "Grouped boxplot with ggplot2"
(https://r-graph-gallery.com/265-grouped-boxplot-with-ggplot2.html).
BRAVE(R) framework: Fox & Boelen, GSLS GenAI Teacher Tutorial, Chapter 3,
CC BY-SA-NC 4.0.*
