---
title: "Translating SPSS Syntax into R with the BRAVE(R) Framework"
author: "Laurence EFrank"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Before you start

Asking an AI tool to translate SPSS syntax can be tricky if you accept the output 
uncritically. The R code will usually look right and run without error, 
but might use a different default than SPSS (a different variance assumption, 
a different skewness formula, a different way of handling missing data). 
The numbers then disagree, and you have no idea why.

So in this exercise AI is a translator you supervise:

1. Give the AI a precise, well-scoped prompt using the **BRAVE(R)** framework.
2. Read the translation and identify where R's defaults differ from SPSS.
3. Run the code on data you control and verify the results make sense.

> **Data safety reminder.** Never paste real, unpublished, or participant data
> into a GenAI tool. This exercise uses **simulated data** you generate yourself,
> so nothing sensitive ever leaves your machine.

## Learning objectives

- Write a BRAVE(R) prompt that asks AI to translate SPSS syntax into R.
- Recognise common SPSS versus R default mismatches and check for them.
- Run and verify an AI-produced translation against simulated data.

---

# Set up your R environment

You only need to do **one** of the two options below.

### Option A: You already have RStudio installed

1. Open **RStudio**.
2. Create a new R Markdown file:
   `File ▸ New File ▸ R Markdown…`, then keep the HTML output and click **OK**.
3. To run a code chunk, click the small green ▶ ("Run Current Chunk") at the
   top-right of the chunk, or place your cursor in it and press
   **Ctrl+Shift+Enter** (Windows/Linux) / **Cmd+Shift+Enter** (Mac).
4. To produce the full HTML document, click **Knit** at the top of the editor.

### Option B: You do not have RStudio: use Posit Cloud (in your browser)

[Posit Cloud](https://posit.cloud) runs RStudio in the browser — nothing to install.

1. Go to **https://posit.cloud** and click **Sign Up** (the free plan is enough).
2. After logging in, click **New Project ▸ New RStudio Project**. Wait a few
   seconds for the workspace to load. You now have the full RStudio interface.
3. Open an R Markwon file: in the **Files** pane (bottom-left) click **File**,
   choose `New File`, then  `New R Markdown File`.
4. Run chunks with the green ▶ button and **Knit** exactly as in Option A.

**Tip.** On the free Posit Cloud plan, projects have limited monthly hours.
Close the project tab when you are done so you don't burn through your time.

---

# The SPSS syntax you will translate

This is the original analysis, written in SPSS syntax. Your job is not to run
it in SPSS, but to obtain and then critically check an R translation of it.

```text
* Descriptive statistics for key variables.
DESCRIPTIVES VARIABLES=score anxiety depression wellbeing
  /STATISTICS=MEAN STDDEV MIN MAX KURTOSIS SKEWNESS.

* Independent samples t-test (group: 1=control, 2=treatment).
T-TEST GROUPS=group(1 2)
  /MISSING=ANALYSIS
  /VARIABLES=score
  /CRITERIA=CI(.95).

* One-way ANOVA with post-hoc tests.
ONEWAY score BY group
  /STATISTICS DESCRIPTIVES HOMOGENEITY
  /MISSING ANALYSIS
  /POSTHOC=TUKEY ALPHA(0.05).

* Pearson correlations with pairwise deletion.
CORRELATIONS
  /VARIABLES=score anxiety depression wellbeing
  /PRINT=TWOTAIL SIG
  /MISSING=PAIRWISE.
```

---

# Step 1 Create the simulated data

This gives you a dataset with exactly the variables the SPSS syntax expects:
`group` (coded **1 = control, 2 = treatment**, just like SPSS), plus `score`,
`anxiety`, `depression`, and `wellbeing`. A few values are set to missing on
purpose, so the `/MISSING` subcommands actually matter.

Copy past the entire R code chunck into your R Markdown file.

```{r simulate-data}
set.seed(2025)

n <- 60                          # respondents per group
group <- rep(c(1, 2), each = n)  # SPSS-style numeric codes: 1 = control, 2 = treatment
treat <- group - 1               # 0 = control, 1 = treatment (helper for the simulation)

# Correlated outcomes with a small treatment effect baked in.
anxiety    <- rnorm(2 * n, mean = 50 - 5 * treat, sd = 10)
depression <- 0.6 * anxiety + rnorm(2 * n, mean = 10, sd = 8)
wellbeing  <- 60 + 6 * treat - 0.3 * anxiety + rnorm(2 * n, mean = 0, sd = 8)
score      <- 100 + 8 * treat - 0.2 * depression + rnorm(2 * n, mean = 0, sd = 15)

dat <- data.frame(group, score, anxiety, depression, wellbeing)

# Inject a few missing values so /MISSING handling becomes visible.
set.seed(99)
dat$anxiety[sample(nrow(dat), 5)]   <- NA
dat$wellbeing[sample(nrow(dat), 4)] <- NA

# Quick look
str(dat)
colSums(is.na(dat))   # how many missing per variable?
```

> **Note on group coding.** SPSS happily uses the numeric codes `1`/`2`. In R you
> will almost always want `group` as a **factor** with readable labels — this is
> itself part of a faithful translation, and the AI may or may not do it for you.

---

# Step 2 Ask AI to translate, using BRAVE(R)

Write a single, well-scoped prompt. The bracketed letters show the BRAVE(R)
component each part covers. Please do not use the BRAVER code in your prompt and
leave out (Expectations), (Role), etc.

Please translate the SPSS syntax below into equivalent R code **(Expectations)**.
Act as a statistician fluent in both SPSS and R **(Role)**,
writing for a social-science researcher who knows SPSS well but is new to R
**(Audience)**. My data frame is called `dat` and contains `group`
(1 = control, 2 = treatment), `score`, `anxiety`, `depression`, and `wellbeing`
**(Variables)**. For **each** SPSS command, give the R equivalent, name the
package you use, and add a one-line comment wherever R's default differs from
SPSS. For example variance assumptions in the t-test, the skewness/kurtosis
formula, p-value adjustment in correlations, and missing-data handling **(Variables)**.
Only produce code I can run and check myself and do not fabricate any results
**(Boundaries)**.
>
> ```
> [paste the entire SPSS syntax here]
> ```

### Refine 

After you read the answer, push on the weak spots. This could be a Refine step:

The SPSS T-TEST output shows both an "equal variances assumed" and an "equal
variances not assumed" row. Which one does your `t.test()` call give by default,
and how do I get the other? **(Refine)**

For the correlations, does your code adjust the p-values for multiple
comparisons? SPSS does not. Please make sure the R version matches SPSS.
**(Refine)**

---

# Step 3 Human oversight: the mismatches to check

Before you trust the translation, walk this checklist. These are the points where
a SPSS to R translation most often goes wrong:

| SPSS behaviour | R default to watch for | What to verify |
|---|---|---|
| **Skewness / kurtosis** use the G1/G2 (SAS/SPSS) formulas | `psych::describe()` uses a different formula unless you set `type = 2` | Did the translation set `type = 2`? |
| **T-TEST** reports *both* Student and Welch, plus Levene's test | `t.test()` gives **Welch only** (`var.equal = FALSE`) | Is the variance assumption stated? Is Levene's test included? |
| **ONEWAY HOMOGENEITY** = Levene's test, shown automatically | R needs `car::leveneTest()` explicitly | Is homogeneity actually tested, not skipped? |
| **CORRELATIONS** print raw two-tailed p-values (no correction) | `psych::corr.test()` adjusts p-values (Holm) unless `adjust = "none"` | Are the p-values uncorrected, like SPSS? |
| **/MISSING=PAIRWISE** vs **ANALYSIS** | R functions differ: `cor()` needs `use = "pairwise.complete.obs"` | Is the deletion method correct per command? |
| `group` is numeric `1`/`2` | Many R model functions expect a **factor** | Was `group` converted to a labelled factor? |

**The oversight habit.** "It ran without error" is not "it matches SPSS." For
each command, ask: would this give the same numbers as SPSS, and if not, why?

---

# Step 4 Run and verify

Paste the AI's translation into the chunk below, run it, and check it against the
checklist above and the model answer in the Appendix. Fix anything that doesn't
match SPSS's behaviour.

```{r your-translation, eval=FALSE}
# >>> YOUR TASK: paste the AI's R translation here, then run and verify it <<<


```

A couple of quick independent checks you can always do yourself:

```{r sanity-checks, eval=FALSE}
# With only TWO groups, a one-way ANOVA F should equal the (Student) t squared.
# This is a verification that your t-test and ANOVA translations agree.
t_out <- t.test(score ~ group, data = dat, var.equal = TRUE)
aov_out <- summary(aov(score ~ factor(group), data = dat))
t_out$statistic^2        # t^2
aov_out[[1]]$`F value`   # ANOVA F  -> should match t^2
```

---

# Step 5 Reflection

1. Which checklist item, if missed, would most change your conclusions — and why?
2. Where did the AI's first translation differ from what SPSS actually does?
3. How did the **Refine** step improve the translation compared with accepting
   the first answer?

---

*BRAVE(R) framework: Fox & Boelen, GSLS GenAI Teacher Tutorial, Chapter 3,
CC BY-SA-NC 4.0.*
